High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-08-09 notebooks/fmnist.qmd
now some progress on fmnist as well

```{julia}
# Search:
generic_generator = WachterGenerator()
ce_wachter = generate_counterfactual(
    x_factual, target, counterfactual_data, M, generic_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
greedy_generator = GreedyGenerator(η=2.0)
ce_jsma = generate_counterfactual(
    x_factual, target, counterfactual_data, M, greedy_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
)
```

```{julia}
p1 = Plots.plot(
    convert2image(FashionMNIST, reshape(x_factual,28,28)),
    axis=nothing, 
    size=(img_height, img_height),
    title="Factual"
)
plts = [p1]

ces = zip([ce_wachter,ce_jsma])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Wachter","JSMA"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(FashionMNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "fashion_you_may_not_like_it.png"))
```

#### REVISE

```{julia}
using CounterfactualExplanations.Models: load_fashion_mnist_vae
vae = load_fashion_mnist_vae()
vae_weak = load_fashion_mnist_vae(;strong=false)
Serialization.serialize(joinpath(output_path,"fashion_mnist_classifier.jls"), M)
Serialization.serialize(joinpath(output_path,"fashion_mnist_vae.jls"), vae)
Serialization.serialize(joinpath(output_path,"fashion_mnist_vae_weak.jls"), vae_weak)
```

```{julia}
# Define generator:
revise_generator = REVISEGenerator(
    opt = Flux.Optimise.Descent(0.25),
    λ=0.0,
)
# Generate recourse:
counterfactual_data.generative_model = vae # assign generative model
ce_strong = generate_counterfactual(
    x_factual, target, counterfactual_data, M, revise_generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
counterfactual_data_weak = deepcopy(counterfactual_data)
counterfactual_data_weak.generative_model = vae_weak
ce_weak = generate_counterfactual(
    x_factual, target, counterfactual_data_weak, M, revise_generator;
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)
```

```{julia}
ces = zip([ce_strong,ce_weak])
counterfactuals = reduce((x,y)->cat(x,y,dims=3),map(ce -> CounterfactualExplanations.counterfactual(ce[1]), ces))
phat = reduce((x,y) -> cat(x,y,dims=3), map(ce -> target_probs(ce[1]), ces))
plts = [p1]
for x in zip(eachslice(counterfactuals; dims=3), eachslice(phat; dims=3), ["Strong VAE","Weak VAE"])
    ce, _phat, _name = (x[1],x[2],x[3])
    _title = "$(_name) (p=$(round(_phat[1]; digits=2)))"
    plt = Plots.plot(
        convert2image(FashionMNIST, reshape(ce,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "fashion_surrogate_gone_wrong.png"))
```

@@

```{julia}
# Hyper:
_retrain = true
_regen = true

# Data:
n_obs = 1000
counterfactual_data = load_fashion_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
counterfactual_data.generative_model = vae
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 150
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 512
activation = Flux.swish
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    Dense(n_hidden, n_hidden, activation),
    Dense(n_hidden, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                   # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

```{julia}
# JEM parameters:
𝒟x = Uniform(-1.0,1.0)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10,
)
α = [1.0,1.0,5e-2]      # penalty strengths
```

@@

# Joint Energy Model:
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    epochs=epochs,
    batch_size=batch_size,
    finaliser=_finaliser,
    loss=_loss,
    jem_training_params=(
        α=α,verbosity=10,
    ),
    sampling_steps=35,
)

# JEM with adversarial training:
jem_adv = deepcopy(jem)
# jem_adv.adv_training = true

# Deep Ensemble of Joint Energy Models:
jem_ens = EnsembleModel(model=jem, n=n_ens)

# Deep Ensemble of Joint Energy Models with adversarial training:
# jem_ens_plus = EnsembleModel(model=jem_adv, n=n_ens)

# Dictionary of models:
models = Dict(
    "MLP" => mlp,
    "MLP Ensemble" => mlp_ens,
    "JEM" => jem,
    "JEM Ensemble" => jem_ens,
    # "JEM Ensemble+" => jem_ens_plus,
)

@@

        _w = 1500
        plts = []
        neach = 10
        for i in 1:10
            x = sampler(f, opt; niter=n_regen, n_samples=neach, y=i)
            plts_i = []
            for j in 1:size(x, 2)
                xj = x[:,j]
                xj = reshape(xj, (n_digits, n_digits))
                plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
            end
            plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
            plts = [plts..., plt]
        end
        plt = Plots.plot(plts..., size=(_w,_w), layout=(10,1), plot_title=mod_name)
        savefig(plt, joinpath(output_images_path, "fashion_mnist_generated_$(mod_name).png"))
        display(plt)
    end
end
```

@@

### Different Models

```{julia}
function _plot_eccco_fashion_mnist(
    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
    λ=[0.1,0.3,0.3],
    temp=0.1,
    plt_order = ["MLP", "MLP Ensemble", "JEM", "JEM Ensemble"],
    opt = nothing,
    rng::Union{Int,AbstractRNG}=1234,
    T::Int = 100,
)

    # Setup:
    Random.seed!(rng)
    if x isa Int
        x = reshape(counterfactual_data.X[:,rand(findall(labels.==x))],input_dim,1)
    end

    # Generate counterfactuals using ECCCo generator:
    eccco_generator = ECCCoGenerator(
        λ=λ, 
        temp=temp, 
        opt=opt,
    )

    ces = Dict()
    for (mod_name, mod) in model_dict
        ce = generate_counterfactual(
            x, target, counterfactual_data, mod, eccco_generator; 
            decision_threshold=γ, max_iter=T,
            initialization=:identity,
            converge_when=:generator_conditions,
        )
        ces[mod_name] = ce
    end
    _plt_order = map(x -> findall(collect(keys(model_dict)) .== x)[1], plt_order)

    # Plot:
    p1 = Plots.plot(
        convert2image(FashionMNIST, reshape(x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title="Factual"
    )

    plts = []
    for (_name,ce) in ces
        _x = CounterfactualExplanations.counterfactual(ce)
        _phat = target_probs(ce)
        _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
        plt = Plots.plot(
            convert2image(FashionMNIST, reshape(_x,28,28)),
            axis=nothing, 
            size=(img_height, img_height),
            title=_title
        )
        plts = [plts..., plt]
    end
    plts = plts[_plt_order]
    plts = [p1, plts...]
    plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))

    return plt, eccco_generator
end
```

```{julia}
plt, eccco_generator = _plot_eccco_fashion_mnist()
display(plt)
savefig(plt, joinpath(output_images_path, "fashion_mnist_eccco.png"))
```

### All digits

```{julia}
function plot_fashion_mnist(
    factual::Int,target::Int;
    generator::AbstractGenerator,
    model::AbstractFittedModel=model_dict["JEM Ensemble"],
    data::CounterfactualData=counterfactual_data,
    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
    _plot_title::Bool=true,
    show_factual::Bool=false,
    kwargs...,
)
    Random.seed!(rng)

    decision_threshold = !isdefined(kwargs, :decision_threshold) ? 0.9 : decision_threshold
    max_iter = !isdefined(kwargs, :max_iter) ? 100 : max_iter
    initialization = !isdefined(kwargs, :initialization) ? :identity : initialization
    converge_when = !isdefined(kwargs, :converge_when) ? :generator_conditions : converge_when

    x = reshape(data.X[:,rand(findall(predict_label(model, data).==factual))],input_dim,1)
    ce = generate_counterfactual(
        x, target, data, model, generator; 
        decision_threshold=decision_threshold, max_iter=max_iter,
        initialization=initialization,
        converge_when=converge_when,
        kwargs...
    )

    _title = _plot_title ? "$(factual) -> $(target)" : ""

    _x = CounterfactualExplanations.counterfactual(ce)
    plt = Plots.plot(
        convert2image(FashionMNIST, reshape(_x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    if show_factual
        plt_factual = Plots.plot(
            convert2image(FashionMNIST, reshape(x,28,28)),
            axis=nothing, 
            size=(img_height, img_height),
            title="Factual"
        )
        plt = Plots.plot(plt_factual, plt; size=(img_height*2,img_height), layout=(1,2))
    end

    return plt
end
```

```{julia}
if _regen
    function plot_all_digits(rng=123;verbose=true,kwargs...)
        plts = []
        for i in 0:9
            for j in 0:9
                @info "Generating counterfactual for $(i) -> $(j)"
                plt = plot_fashion_mnist(i,j;kwargs...,rng=rng)
                !verbose || display(plt)
                plts = [plts..., plt]
            end
        end
        plt = Plots.plot(plts...; size=(img_height*10,img_height*10), layout=(10,10))
        return plt
    end
    plt = plot_all_digits(generator=eccco_generator)
    savefig(plt, joinpath(output_images_path, "fashion_mnist_eccco_all_digits.png"))
end
```

## Benchmark

```{julia}
λ₁ = 0.1
λ₂ = 0.3
λ₃ = 0.3
Λ = [λ₁, λ₂, λ₃]

# Benchmark generators:
generator_dict = Dict(
    "Wachter" => WachterGenerator(λ=λ₁),
    "REVISE" => REVISEGenerator(λ=λ₁),
    "Schut" => GreedyGenerator(),
    "ECCCo" => ECCCoGenerator(λ=Λ),
)
```

```{julia}
# Measures:
measures = [
    CounterfactualExplanations.distance,
    ECCCo.distance_from_energy,
    ECCCo.distance_from_targets,
    CounterfactualExplanations.Evaluation.validity,
    CounterfactualExplanations.Evaluation.redundancy,
    ECCCo.set_size_penalty
]

bmks = []
for target in sort(unique(labels))
    for factual in sort(unique(labels))
        if factual == target
            continue
        end
        bmk = benchmark(
            counterfactual_data; 
            models=model_dict, 
            generators=generator_dict, 
            measure=measures,
            suppress_training=true, dataname="FashionMNIST",
            n_individuals=5,
            target=target, factual=factual,
            initialization=:identity,
            converge_when=:generator_conditions,
        )
        push!(bmks, bmk)
    end
end
bmk = reduce(vcat, bmks)

CSV.write(joinpath(output_path, "fashion_mnist_benchmark.csv"), bmk())
```

```{julia}
df = @chain bmk() begin
    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
    @mutate(variable = ifelse.(variable .== "redundancy", "Redundancy", variable))
    @mutate(variable = ifelse.(variable .== "Validity", "Validity", variable))
end
plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
plt = draw(
    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
    facet=(; linkyaxes=:minimal)
)   
display(plt)
save(joinpath(output_images_path, "fashion_mnist_benchmark.png"), plt, px_per_unit=5)
```

>>>>

@@

```{julia}
# Hyper:
_retrain = false
_regen = false

# Data:
n_obs = 10000
counterfactual_data = load_fashion_mnist(n_obs)
counterfactual_data.X = pre_process.(counterfactual_data.X)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
x_factual = reshape(pre_process(x_factual, noise=0.0f0), input_dim, 1)
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 10
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 512
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden, activation),
    # Dense(n_hidden, n_hidden, activation),
    # Dense(n_hidden, n_hidden, activation),
    Dense(n_hidden, n_out),
)
n_ens = 5                                  # number of models in ensemble
_loss = Flux.Losses.logitcrossentropy       # loss function
_finaliser = x -> x                         # finaliser function
```

@@

# Dictionary of models:
models = Dict(
    # "MLP" => mlp,
    "Ensemble" => mlp_ens,
)

@@

        _w = 1000
        plts = []
        neach = 1
        for i in 1:10
            x = sampler(f, opt; niter=n_regen, n_samples=neach, y=i)
            plts_i = []
            for j in 1:size(x, 2)
                xj = x[:,j]
                xj = reshape(xj, (n_digits, n_digits))
                plts_i = [plts_i..., Plots.heatmap(rotl90(xj), axis=nothing, cb=false)]
            end
            plt = Plots.plot(plts_i..., size=(_w,0.10*_w), layout=(1,10))
            plts = [plts..., plt]
        end
        plt = Plots.plot(plts..., size=(_w,_w/10), layout=(1,10), plot_title=mod_name)
        savefig(plt, joinpath(output_images_path, "fashion_mnist_generated_$(mod_name).png"))
        display(plt)
    end
end
```

@@

```{julia}
function plot_fmnist(
    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
    generator = ECCCoGenerator(),
    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
    T::Int = 100,
    use_class_loss::Bool = true,
    model=model_dict["Ensemble"],
    img_height::Int = img_height,
    test_data::Bool = false,
    kwrgs...,
)

    # Setup:
    Random.seed!(rng)
    if x isa Int
        x_fact = counterfactual_data.X[:,rand(findall(labels.==x))][:,:]
    else
        x_fact = x
    end

    if test_data
        data = load_mnist_test()
    else
        data = counterfactual_data
    end

    ce = generate_counterfactual(
        x_fact, target, data, model, generator; 
        decision_threshold=γ, max_iter=T,
        initialization=:identity,
        converge_when=:generator_conditions,
    )

    # Plot:
    p1 = Plots.plot(
        convert2image(MNIST, reshape(x_fact,28,28)),
        axis=([], false),
        size=(img_height, img_height),
        title="Factual"
    )

    _x = CounterfactualExplanations.counterfactual(ce)
    p2 = Plots.plot(
        convert2image(MNIST, reshape(_x,28,28)),
        axis=([], false), 
        size=(img_height, img_height),
        title="Counterfactual"
    )

    plt = Plots.plot(p1, p2; size=(img_height*2,img_height), layout=(1,2), kwrgs...)

    return plt
end

```

```{julia}
rng = rand(1:10000)
factual = 3
plt = plot_fmnist(3, 1, rng=rng)
```

### 2023-08-09 notebooks/mnist.qmd
rebuttals nearly revised
now some progress on fmnist as well

```{julia}
_plt_order = [
    "MLP", 
    "MLP Ensemble", 
    "Large Ensemble (n=50)", 
    "LeNet-5",
    # "LeNet-5 Ensemble",
    # "RobustNet",
    "JEM", 
    "JEM Ensemble",
]
plt_additional_models, _, _ces_ = _plot_eccco_mnist(
    λ = [0.125,0.25,0.25],
    plt_order = _plt_order,
    model_dict=large_model_dict,
    wide = true,
    plot_factual = true,
    img_height = 150,
)
display(plt_additional_models)
# savefig(plt_additional_models, joinpath(output_images_path, "mnist_eccco_additional.png"))
```

```{julia}
combos = [
    (7,2),
    (0,8),
    (8,3),
    (9,7),
    (4,9),
    (2,1),
    (3,5),
    (5,8),
    (6,0),
    (1,7),
]

plts = []

for (factual, target) in combos
    plt, _, _ = _plot_eccco_mnist(
        factual, target; 
        plt_order = _plt_order,
        model_dict = large_model_dict,
        wide = true,
        plot_factual = true,
        img_height = 150
    )
    display(plt)
    push!(plts, plt)
end
```

```{julia}
λ = [0.1,0.25,0.25]

@@

combos = [
    (9,7),
    # (9,7),
    # (9,7),
    (6,0),
    # (6,0),
    # (6,0),
    (4,9),
    # (4,9),
    # (4,9),
    (8,3),
    # (8,3),
    # (8,3),
    (5,6),
    # (5,6),
    # (5,6),
]

@@

>>>>

```{julia}
_plt_order = [
    "MLP", 
    "MLP Ensemble", 
    "Large Ensemble (n=50)", 
    "LeNet-5",
    # "LeNet-5 Ensemble",
    # "RobustNet",
    "JEM", 
    "JEM Ensemble",
]
plt_additional_models, _, _ces_ = _plot_eccco_mnist(
    λ = [0.1,0.21,0.21],
    plt_order = _plt_order,
    model_dict=large_model_dict,
    wide = true,
    plot_factual = true,
    img_height = 150,
)
display(plt_additional_models)
# savefig(plt_additional_models, joinpath(output_images_path, "mnist_eccco_additional.png"))
```

λ = [0.1,0.215,0.215]

@@

combos = [
    (9,7),
    (6,0),
    (4,1),
    (2,3),
    (5,8),
]

n_each = 2
combos = reduce(vcat, [fill(c, n_each) for c in combos])

@@

```{julia}
final_plts = []
for (i, (factual, target)) in enumerate(combos)
    p1 = plts_eccco[i]
    p2 = plts_wachter[i]
    plt = Plots.plot(
        p1, p2, layout=(2,1), plot_title="Factual: $factual, Target: $target",
        size = (1200, 400),
    )
    display(plt)
    savefig(plt, joinpath("dev/rebuttal/www", "mnist_$(factual)to$(target)_$(i).png"))
    push!(final_plts, plt)
end
```
